\documentclass{article}
\usepackage{url}
\usepackage{named}
\usepackage{yfonts}

\newtheorem{lem}{Lemma}
\newtheorem{thm}{Theorem}
\newtheorem{cor}{Corollary}

\def\ar{\leftarrow}
\def\lrar{\leftrightarrow}
\def\beq{\begin{equation}}
\def\eeq#1{\label{#1}\end{equation}}
\def\ba{\begin{array}}
\def\ea{\end{array}}
\def\gringo{{\sc gringo}}
\def\clingo{{\sc clingo}}
\def\anthem{{\sc anthem}}
\def\vampire{{\sc vampire}}
\def\num{\overline}
\def\p2f{\hbox{p2f}}
\def\no{\emph{not\/}}
\def\head{\emph{Head\/}}
\def\body{\emph{Body\/}}
% \def\uc{\widetilde\forall}
\def\Sfr{\textfrak{S}}
\def\val#1#2{\emph{val\,}_{#1}({#2})}

\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}

\hyphenation{lif-schitz}

\begin{document}

\title{\bf External Behavior of a Logic Program\\
  and Verification of Refactoring}
\author{Jorge Fandinno, Zachary Hansen, Yuliya Lierler\\
  University of Nebraska Omaha\\ \\
Vladimir Lifschitz and Nathan Temple\\ University of Texas at Austin}
\date{}

\maketitle

\begin{center}INCOMPLETE DRAFT\end{center}

\section{Introduction}

This note is about the process of refactoring in the context of answer
set programming, that is, about modifying an ASP program without changing
its external behavior.  Refactoring a program usually involves a series of
small changes that improve its structure or performance.  In answer set
programming, refactoring can serve also for converting code that is
semantically correct but cannot be executed by the existing solvers into an
executable program.  For example, the {\sc clingo} program
\begin{verbatim}
  composite(I*J) :- I > 1, J > 1.
        prime(I) :- I = a..b, not composite(I).
\end{verbatim}
is a semantically correct definition of the set of primes in the
interval $\{a,\dots,b\}$, assuming that $a>1$.  This program
is not executable, because the variables in
the rule defining \verb|composite/1| are unsafe.  The error message
\begin{verbatim}
  grounding stopped because of errors
\end{verbatim}
will not disappear even if we indicate that the predicate
\verb|composite/1| is auxiliary by appending the directive
\begin{verbatim}
  #show prime/1.
\end{verbatim}
A safe program can be obtained by rewriting the first rule as
\begin{verbatim}
  composite(I*J) :- I = 2..b, J = 2..b.
\end{verbatim}
This is an example of refactoring, because this modification
does not change the extent of \verb|prime/1|.
As the next refactoring step, we can optimize the program using the
fact that every composite
number in the interval $\{a,\dots,b\}$ has a divisor in the interval
$\{2,\dots\lfloor \sqrt b\rfloor\}$:
\begin{verbatim}
       sqrt_b(M) :- M = 1..b, M*M <= b, (M+1)*(M+1) > b.
  composite(I*J) :- sqrt_b(M), I = 2..M, J = 2..b.
        prime(I) :- I = a..b, not composite(I).
\end{verbatim}

We are interested in the problem of verifying the correctness of
refactoring steps like these. We would like to make the concept of external
behavior of a logic program precise, and to
design software tools for verifying that a proposed modification of an
ASP program would have no effect on its external behavior.

In the Abstract Gringo language \cite{geb15}, a program is defined as
a set of rules, so that a program
includes neither directives nor comments.  Under this narrow definition,
the program itself does not tell us which predicate symbols are meant to
represent the output, and which symbols are considered
auxiliary.  Changing the
extent of an auxiliary predicate in the process of refactoring, such
as the changes in the extent of \verb|composite/1| above, and
the presence of new auxiliary symbols, such as
\verb|sqrt_b/1|, do not indicate a mistake.
So the difference between output symbols and auxiliary symbols is essential
for the idea of refactoring.

Furthermore, the rules of a program do not show what kind of input is
supposed to be provided for it.
Generally, an input for an ASP program can be specified in two ways.
First, some symbolic constants, such as {\tt a} and {\tt b} in the
programs above, may be meant to serve as placeholders for elements of
the input.
There is no way to decide by looking at the rules
which symbolic constants are supposed to play this role.
Second, some predicate symbols occurring in the program may occur in
the bodies of rules only, not in the heads.  The extents of such
predicates may be specified as part of input when we run the program.

Some inputs may not conform to the programmer's assumptions about
the intended use of the program.  For instance, the program above is
expected to be applied to intervals $\{a,\dots,b\}$ such that $a>1$;
we are not interested in the cases when the placeholder~{\tt a} is
replaced by an integer that is not greated than~1 or by a symbolic constant.
Stable models of a program for ``bad'' inputs are not related to its
external behavior if the program is used as intended.

To sum up, what we consider external behavior of a set of rules
depends on how these rules are meant to be used.

In this paper, we make this idea precise
for the subset of
Abstract Gringo called mini-\gringo\ \cite[Section~2]{fan20},
\cite[Section~2.1]{fan22}.   Then we study the special case when
the proof assistant \anthem\ 
\cite{fan20} and the theorem prover \vampire\ \cite{vor13}
 can be used to verify that two mini-\gringo\ programs have the
 same external behavior.  The three versions of the prime number program
above will be used as a running example.

\section{On the syntax of mini-\gringo}\label{sec:mg}

There are minor syntactic differences between
mini-\gringo\ and the input language of the grounder \gringo,
explained by the fact the former is designed for theoretical studies, and the
latter for actual programming.
For example, the optimized prime number program from the introduction,
rewritten in the syntax of mini-\gringo, becomes
$$\ba {rcl}
\emph{sqrt\_b\/}(M) &\ar& M = \num  1\,..\,b \,\land\, M\times M \leq b
          \,\land\, (M+\num 1)\times(M+\num 1) > b,\\
          \emph{composite\/}(I\times J) &\ar& \emph{sqrt\_b\/}(M) \,\land\,
                 I = \num 2\,..\,N \,\land\, J = \num 2\,..\,b,\\
                 \emph{prime\/}(I) &\ar& I = a\,..\,b
                 \,\land\, \emph{not}\ \emph{composite\/}(I).
\ea
$$
The symbols $\num 1$ and $\num 2$ are ``numerals''---syntactic objects
representing integers. Distinguishing between an integer and the corresponding
numeral is sometimes convenient.  In examples of rules and programs, we will
freely switch between the two styles.

In the definition of mini-\gringo,
\emph{precomputed terms} are numerals,
  symbolic constants, and the symbols \emph{inf}, \emph{sup}.
  We assume that a total order on precomputed terms in chosen, such that
  \emph{inf} is its least element, \emph{sup} is its greatest element, and,
 for all integers~$m$ and~$n$,   $\num m < \num n$ iff $m<n$.
A  \emph{precomputed atom} is an expression of the form $p({\bf t})$, where~$p$
  is a symbolic constant and~$\bf t$ is a tuple of precomputed terms.
  A \emph{predicate symbol} is a pair $p/n$, where~$p$ is a symbolic
  constant and~$n$ is a nonnegative integer.  About an atom $p({\bf t})$ we
  say that it \emph{contains} $p/n$ if the tuple~$\bf t$ consists of~$n$
  terms.

\section{User guides}

{\bf Definition 1}$\;$
A \emph{user guide} is a quadruple
\beq
(\emph{PH},\emph{In},\emph{Out},\emph{Dom\/})
\eeq{ug}
where
\begin{itemize}
\item \emph{PH} is a finite set of symbolic constants, called
  \emph{placeholders},
\item \emph{In} and \emph{Out} are disjoint finite sets of predicate
  symbols, called \emph{input symbols} and \emph{output symbols},
  and
\item \emph{Dom} is a set such that each of its elements is a pair $(v,\I)$,
  where
  \begin{itemize}
    \item [(i)]$v$ is a function that maps the set~\emph{PH} of placeholders
      to the set of precomputed terms that are different from
      placeholders, and
    \item  [(ii)] $\I$ is a subset of the set of precomputed atoms that contain
      an input symbol and do not contain placeholders.
\end{itemize}
\end{itemize}

The set \emph{Dom} is the \emph{domain} of the user guide, and
pairs $(v,\I)$ satisfying conditions~(i) and~(ii) are called \emph{inputs}
\cite[Section~5.2]{fan20}.
An input $(v,\I)$ represents a way to choose the values of
placeholders and the extents of input predicates: for every
placeholder~$c$, specify $v(c)$ as its value, and add the atoms~$\I$
to the rules of the program as facts.  If~$\Pi$ is a mini-\gringo\ program
then $v(\Pi)$ stands for the mini-\gringo\ program obtained from~$\Pi$ by
replacing every occurrence of every constant~$c$ in the domain of~$v$
by~$v(c)$
\cite[Section~2.4]{fan22}.  Using this notation, we can say that
choosing $(v,\I)$ as input for~$\Pi$ amounts to
replacing~$\Pi$ by the program $v(\Pi)\cup\I$.

To use a program in accordance with user guide~(\ref{ug})
means to run it for inputs that belongs to \emph{Dom}.  The
inputs that do not belong to \emph{Dom} are not related to the
external behavior of the program when it is used as intended.

\medskip\noindent{\bf Example 1}$\;$ The intended use of the
programs discussed in the introduction is
described by user guide~(\ref{ug}) with
$$\emph{PH} = \{a,b\},\ \emph{In} = \emptyset,\
  \emph{Out} = \{\emph{prime}/1\},$$
and with the domain consisting of the inputs $(v,\emptyset)$ such that
$v(a)$ and $v(b)$ are numerals, and $v(a)>\num 1$.  We will denote this
user guide by $\emph{UG}_p$.

\medskip\noindent{\bf Example 2}$\;$ We would like to describe the meaning
of the word \emph{orphan} by a logic program \cite[Section ??]{gel14}.
The intended use of such a
program can be described by user guide~(\ref{ug}) with
$$\emph{PH} = \emptyset,\
\emph{In} = \{\emph{father/2},\emph{mother/2}, \emph{living/1}\},\
\emph{Out} = \{\emph{orphan/1}\},$$
and with the domain consisting all inputs.    We will denote this
{ser guide by $\emph{UG}_o$.  We will examine
two possible definitions of \emph{orphan/1} and consider the question of their
equivalence with respect to $\emph{UG}_o$.
\medskip

User guides are closely related to
lp-functions \cite[Section~2]{gel02}
    and to io-programs \cite[Section~5]{fan20}.
    Combining a set of rules with a user guide is similar to using these
    rules as part of
    an lp-function or an io-program.

\section{External behavior}

    An \emph{output atom} of a user guide~\emph{UG} is a precomputed atom
    that contains an output symbol of~\emph{UG}.
    
    \medskip\noindent{\bf Definition 2}$\;$
    Let~$(v,\I)$ be an input in the domain
    of a user guide~\emph{UG}, and let~$\Pi$ be
a mini-\gringo\ program such that the heads of its rules do not
contain input symbols of~\emph{UG}.  The \emph{external behavior} of~$\Pi$
for the user guide~\emph{UG} and the input~$(v,\I)$ is the collection
of all sets that can be represented as the intersection of a stable model
of $v(\Pi)\cup\I$ with the set of output atoms of~\emph{UG}.

\medskip\noindent{\bf Example~1, continued}$\;$ If~$\Pi$ is one of the
three prime number programs from the introduction, and $(v,\I)$ is an
input in the domain of $\emph{UG}_p$, then the program $v(\Pi)\cup\I$
is $v(\Pi)$, and it has a unique stable model.  If~$v$ is defined by
$$v(a)=\num{10},\ v(b)=\num{15},$$
then that stable model includes the atoms $\emph{prime}(\num{11})$,
$\emph{prime}(\num{13})$, and some atoms containing
$\emph{composite}/1$.  The external behavior of each of the programs
for this input is
$$\{\{\emph{prime}(\num{11}),\emph{prime}(\num{13})\}\}.$$
For the safe and optimized versions, this external behavior
can be calculated by instructing \clingo\ to
find all answers for the file obtained from the program by appending the
declarations
\begin{verbatim}
  #const a = 10.
  #const b = 15.
  #show prime/1.
\end{verbatim}

\medskip\noindent{\bf Example~2, continued}$\;$ If~$\Pi$ is the program
\beq\ba l
\verb|parent_living(X) :- father(Y,X), living(Y).|\\
\verb|parent_living(X) :- mother(Y,X), living(Y).|\\
\verb|       orphan(X) :- living(X), not parent_living(X).|
\ea\eeq{orphan1}
and $(v,\I)$ is an input in the domain of $\emph{UG}_o$, then the
program $v(\Pi)\cup\I$ is $\Pi\cup\I$, and it has a unique stable model.
If~$\I$ is
\beq\ba l
\{\emph{father\/}(\emph{jacob},\emph{joseph}),
\emph{mother\/}(\emph{rachel},\emph{joseph}),\\
\quad \quad \emph{living}(\emph{jacob}),
\emph{living}(\emph{rachel}),
\emph{living}(\emph{joseph})\},
\ea\eeq{oi}
then that stable model includes the atoms
$$\emph{orphan\/}(\emph{jacob}),\ \emph{orphan\/}(\emph{rachel})$$
and some atoms containing predicate symbols other than \emph{orphan/1}.
The external behavior of this program for $\emph{UG}_o$ and input~(\ref{oi})
is
\beq
\{\{\emph{orphan\/}(\emph{jacob}),\emph{orphan\/}(\emph{rachel})\}\}.
\eeq{ebor}
It can be calculated by instructing \clingo\ to
find all answers for the file obtained from the program by appending the
facts
\begin{verbatim}
  father(jacob,joseph).
  mother(rachel,joseph).
  living(jacob; rachel; joseph).
\end{verbatim}
and the directive
\begin{verbatim}
  #show orphan/1.
\end{verbatim}

\medskip\noindent{\bf Example~3}$\;$
Assume that \emph{UG} has no placeholders and no input symbols,
that its set of output symbols includes all predicate symbols occurring
in~$\Pi$, and that its domain is $\{(\emptyset,\emptyset)\}$.  Then
the external
behavior of~$\Pi$ with respect to~\emph{UG} and $(\emptyset,\emptyset)$
is the set of stable models of~$\Pi$.

\section{Equivalence}

\noindent{\bf Definition 3}$\;$
Let~\emph{UG} be a user guide, and let~$\Pi_1$, $\Pi_2$ be mini-\gringo\
programs such that the heads of their rules do not contain input symbols
of~\emph{UG}.  We say that~$\Pi_1$ is \emph{equivalent to~$\Pi_2$ with
respect to}~\emph{UG} if, for every input $(v,\I)$ in the domain of~\emph{UG},
the external behavior of $\Pi_1$ for~\emph{UG} and~$(v,\I)$ is the
same as the external behavior of~$\Pi_2$.

\medskip\noindent{\bf Example~1, continued}$\;$ The three programs from the
introduction are equivalent to each other with respect to $\emph{UG}_p$.
We will see that this claim can be verified using the automated reasoning
tools \anthem\ and \vampire.

\medskip\noindent{\bf Example~2, continued}$\;$ Perhaps surprisingly,
the one-rule program
\beq\ba l
\verb|orphan(X) :- living(X), father(Y,X), mother(Z,X),|\\
\verb|             not living(Y), not living(Z).|
\ea\eeq{orphan2}
is not equivalent to~(\ref{orphan1}) with respect to $\emph{UG}_o$.
Indeed, the external behavior of this program with respect
to $\emph{UG}_o$ and input~(\ref{oi}) is $\{\emptyset\}$, which
is different from~(\ref{ebor}).  We will see
that \anthem\ and \vampire\ can help us clarify the relationship between
programs~(\ref{orphan1}) and~(\ref{orphan2}).

\medskip\noindent{\bf Example~3, continued}$\;$
Assume that \emph{UG} has no placeholders and no input symbols,
that its set of output symbols includes all predicate symbols occurring
in~$\Pi_1$ and $\Pi_2$, and that its domain is $\{(\emptyset,\emptyset)\}$.
Then~$\Pi_1$ and~$\Pi_2$ are equivalent to each other with
respect to~\emph{UG} iff they have the same stable models.

\medskip
We understand refactoring a mini-\gringo\ program with respect to a
  user guide~\emph{UG} as replacing it by a program that is equivalent
  to it with respect to~\emph{UG}.

This equivalence relation is essentially an
example of relativized uniform equivalence with projection
\cite{oet08}, except that the language discussed in that paper includes
neither arithmetic operations nor placeholders.
It is \emph{uniform} equivalence, because the programs are extended by
adding facts, rather than more complex rules; \emph{relativized},
because these
  facts~$\I$ are assumed to be atoms containing input symbols,
  not arbitrary atoms; \emph{with projection}, because
  we look at the output atoms in the stable model, not the entire model.

\section{Formal notation for user guides}

To design software for verifying the equivalence of
programs with respect to a user guide, we need to
represent user guides in formal notation.   The format that we chose for
user guide (\verb|.ug|) files is similar to the format of
specification (\verb|.spec|)
files that \anthem\ operates with \cite[Section~7]{fan20}.  Placeholders
and input symbols are represented in a \verb|.ug| file by
\verb|input| statements, for instance:
\begin{verbatim}
  input: n.
  input: living/1, father/2, mother/2.
\end{verbatim}
Output symbols are represented by \verb|output| statements:
\begin{verbatim}
  output: prime/1.
\end{verbatim}
There can be several statements of both kinds in a user guide file, in any
order.

The question of representing the domain \emph{Dom} by a string of
characters is more difficult, because the domain is a set of inputs,
which is generally infinite.
One approach is to define \emph{assumptions} as sentences of an
appropriate first-order language, and characterize the domain by a
list of assumptions; an input belongs to the domain iff it
satisfies all assumptions in that set.  In a user guide file, every
assumption will be represented by a statement similar to \verb|assume|
statements found in \anthem\ specification files.

The language of assumptions for a set \emph{In} of predicate symbols
is a first-order language
with two sorts: the sort \emph{generic} and its subsort
\emph{integer}.
Variables of the first sort are meant to
range over arbitrary precomputed terms, and we will identify them with
variables used in mini-\gringo\ rules.  Variables of the second sort are meant
to range over numerals (or, equivalently, integers).\footnote{The need to
  use a language with two sorts is explained by the fact that function symbols
in a first-order language are supposed to represent total functions, and
arithmetic operations are not defined on symbolic constants.}
The signature $\sigma(\emph{In})$ of this language includes
\begin{itemize}
\item all precomputed terms as object constants; an object constant
  is assigned the sort \emph{integer} iff it is a numeral;
\item the symbols~$+$, $-$ and~$\times$ as binary function constants;
  their arguments and values have the sort \emph{integer};
\item all predicate symbols~$p/n$ from \emph{In} as $n$-ary predicate
  constants; their arguments have the sort \emph{general};
\item the comparison symbols
$$\neq\quad<\quad>\quad\leq\quad\geq$$
as binary predicate constants; their arguments have the sort \emph{general}.
\end{itemize}
(Equality is not mentioned here because it is considered part of any
first-order language.)  An atomic formula $(p/n)({\bf t})$ can be abbreviated
as $p({\bf t})$. An atomic formula $\prec\!\!(t_1,t_2)$, where~$\prec$ is a
comparison symbol, can be written as $t_1\prec t_2$.

We follow the convention adopted in {\sc anthem\/}:
general variables start with $U$, $V$, $W$, $X$, $Y$, and $Z$;
integer variables start with $I$, $J$, $K$, $L$, $M$, and $N$.
For example, the formula
$\exists X (N=X)$ expresses that the value of~$N$ is an object of the sort
\emph{general\/};  it is universally true, because \emph{integer} is a
subsort of \emph{general}.  The formula
$\exists N (N=X)$ expresses that the value of~$X$ is an object of the sort
\emph{integer\/}; it is generally not true.

To use sentences over the signature $\sigma(\emph{In})$ as conditions
on an input, we need to relate inputs to interpretations of this
signature in the sense of first-orger logic.  For every input $(v,\I)$, by
$I(v,\I)$ we denote the interpretation of~$\sigma(\emph{In})$
defined as follows:
\begin{itemize}
\item
the domain of the sort \emph{general} in~$I(v,\I)$
  is the set of all precomputed terms;
\item
the domain of the sort \emph{integer} in~$I(v,\I)$ is the set of all numerals;
\item $I(v,\I)$ interprets every precomputed term~$t$ as~$v(t)$ if~$t$
  is a symbolic constant in the domain of~$v$, and as~$t$ otherwise;
\item $I(v,\I)$ interprets the symbols for arithmetic operations
  as usual in arithmetic;
\item if $p/n$ is a predicate constant from~\emph{In}, and $\bf c$
  is an $n$-tuple of object constants, then $I(v,\I)$ interprets
  $p({\bf c})$ as true iff $p({\bf c})\in\I$;
\item $I(v,\I)$ interprets the comparison symbols as in the definition of
  mini-\gringo\ (see Section~\ref{sec:mg}).
\end{itemize}

Besides \verb|input| and \verb|output| statements, a user guide
file may include one or more statements consisting of the word
\verb|assume| followed by a list of sentences over the
signature~$\sigma(\emph{UG\/})$, called \emph{assumptions}.
The domain of the user guide is the set of inputs $(v,\I)$ such that
the interpretation $I(v,\I)$ satisfies all its assumptions.

\medskip\noindent{\bf Example~1, continued}$\;$ The user guide $\emph{UG}_p$
can be described by the statements
\begin{verbatim}
  input: a, b.
  assume: exists N (a = N) and exists N (b = N).
  assume: a > 1.
  output: prime/1.
\end{verbatim}
The first two lines can be written be more concisely as
\begin{verbatim}
  input: a -> integer, b -> integer.
\end{verbatim}

\medskip\noindent{\bf Example~2, continued}$\;$ The user guide $\emph{UG}_o$
can be described by the statements
\beq\ba l
\verb| input: living/1, father/2, mother/2.|\\
\verb| output: orphan/1.|
\ea\eeq{ugo}
The absence of \verb|assume| statements shows that the domain of this
user guide is the set of all inputs.

\section{Proof assistant {\sc anthem-p2p}}

{\sc anthem-p2p} is an extension of the proof assistant \anthem\
\cite{fan20}.  It uses the resolution theorem prover \vampire\
\cite{vor13} for verifying equivalence of mini-\gringo\ programs with
respect to a user guide.

For example, to verify that first two versions of the prime number program
from the introduction are equivalent with respect to the user guide
$\emph{UG}_p$, we run {\sc anthem-p2p} on the following input files: the
unsafe program
\begin{verbatim}
  composite(I*J) :- I > 1, J > 1.
        prime(I) :- I = a..b, not composite(I).
\end{verbatim}
the safe program
\begin{verbatim}
  composite(I*J) :- I = 2..b, J = 2..b.
        prime(I) :- I = a..b, not composite(I).
\end{verbatim}
and the user guide
\begin{verbatim}
  input: a -> integer, b -> integer.
  assume: a > 1.
  output: prime/1.
\end{verbatim}
{\sc anthem-p2p} transforms the task of verifying equivalence
with respect to a user guide into the problem of verifying the provability of
a formula in some
first-order theory over the signature~$\sigma(\emph{In})$, and
submits that problem to \vampire.  This transformation is discussed below in
Section~??.

The user can help \vampire\ organize search process more efficiently by
supplying {\sc anthem-2p2} with ``helper'' files.  Such a file may instruct
\vampire\ to prove a series of lemmas before trying to prove the goal formula.
A helper file can suggest also instances of the induction schema that may be
useful for the job in hand.

The use of {\sc anthem-2p2} for proving
equivalence of programs is, generally, an interactive process.
If \vampire\ does not prove the goal formula in the allotted time then one of
the options is to give \vampire\ more lemmas to work on and run it
again.  Alternatively, the user can look for a counterexample that refutes the
equivalence claim, as in Example~2 above.  {\sc anthem-2p2} can possibly
help us clarify the source of a puzzling discrepancy between two versions
of a program if we run it in the presence of additional assume statements.  If
adding an assumption to the user guide makes the programs equivalent then
it is possible that perceiving that assumption as self-evident
is the reason why the discrepancy is puzzling.  For
instance, we can observe that
the {\sc anthem-2p2}/\vampire\ team
proves the equivalence of
program~(\ref{orphan1}) to program~(\ref{orphan2}) if we extend user
guide~(\ref{ugo}) by two existence and uniqueness assumptions:
\begin{verbatim}
  assume: forall X exists Y forall Z (father(Z,X) <-> Y=Z).
  assume: forall X exists Y forall Z (mother(Z,X) <-> Y=Z).
\end{verbatim}

\section{Review: second-order completion}

Let \emph{In} and \emph{Out} be disjoint sets of predicate symbols.
Considera mini-\gringo\ program~$\Pi$ such that predicate
symbols from \emph{In} do not occur in the heads of its rules.
If a predicate symbol~$p/n$ belongs neither to~\emph{In}
nor to~\emph{Out} but is contained in an atom that occurs in a rule of~$\Pi$
then we say~$p/n$ is a \emph{private symbol} of~$\Pi$.  The
\emph{first-order completion} of~$\Pi$ is the conjunction of completed
definitions of the symbols from \emph{Out} and the private symbols of~$\Pi$.
The \emph{second-order completion} of~$\Pi$ is the formula obtained from
the first-order completion of~$\Pi$ by replacing all private symbols by
existentially quantified predicate variables
\cite[Section~2.5].\footnote{To be precise, the definition of
  completion in that paper refers to a
  ``program with input and output,'' which has four
  components:~$\Pi$, \emph{In}, \emph{Out}, and a set \emph{PH} of
  symbolic constants.  However, the first-order and second-order
  completion formulas do not depend on the choice of~\emph{PH}.  Second,
  the brief summary above is only applicable to programs without constraints.}

If, for instance, $\emph{In}=\emptyset$, $\emph{Out}=\{q/2\}$, and~$\Pi$
is the program
$$
\ba l
p(a),\\
p(b),\\
\{q(X,Y)\} \ar p(X) \land p(Y),
\ea$$
then the first-order completion of~$\Pi$ is
definitions of~$p/1$ and~$q/2$ are
$$\ba l
\forall V(p(V) \lrar V=a \lor V=b)\\
\land\,
\forall V_1V_2(q(V_1,V_2) \lrar \exists XY(q(V_1,V_2) \land p(X) \land p(Y)
\land V_1=X \land V_2=Y)),
\ea$$
and the second-order completion is
\beq
\ba l
\exists P(\forall V(P(V) \lrar V=a \lor V=b)\,\land\\
\quad\;\;\;\forall V_1V_2(q(V_1,V_2) \lrar \exists XY(q(V_1,V_2) \land
P(X) \land P(Y) \land V_1=X \land V_2=Y))).
\ea
\eeq{sof}

Second-order quantifiers in completion formulas can be eliminated if~$\Pi$
is a ``program without private recursion''
\cite[Section~6.4]{fan20}.  For instance, formula~(\ref{sof}) is equivalent
to the first-order formula
$$
\forall V_1V_2(q(V_1,V_2) \to (V_1=a \lor V_1=b)\land(V_2=a \lor V_2=b)).
$$

\section*{Acknowledgements}

Thanks to Michael Gelfond for comments on a draft of this paper.

\bibliographystyle{named}
\bibliography{bib}
\end{document}

% \medskip
% In the rest of the paper, the domain of a user guide is supposed to be
% specified by a list of assumptions, as in the examples above.


\section{Equivalence of tight programs}

Recall that an \emph{io-program} is a quadruple
\beq
(\Pi,\emph{PH},\emph{In},\emph{Out\/}),
\eeq{iop}

The definition of a tight io-program  \cite[Section~6.2]{fan20}
refers to only the first of its
four components, which is a mini-\gringo\ program.  In view
of this fact, we can talk about tightness of mini-\gringo\ programs.

An interpretation~$I$ of~$\sigma(\emph{In})$ is \emph{standard outside of} a
set \emph{PH} of
symbolic constants if it is either standard\footnote{Standard interpretations
  are defined in the appendix.} or can be obtained from a standard
interpretation by changing the way it interprets symbols from~\emph{PH}
\cite[Section~5]{fan22}.

By slight abuse of notation, in the statement of the theorem below we
write \emph{Asm} for the conjunction of all sentences in \emph{Asm}.

\begin{thm}\label{thm1}
Let~UG be a user guide (PH,In,Out,Asm).
For any tight mini-\gringo\ programs $\Pi_1$,~$\Pi_2$ that do
not contain the input symbols of~UG in the heads of rules,
$\Pi_1$ is equivalent to~$\Pi_2$ with respect to UG iff the sentence
\end{thm}
\beq
\emph{Asm}\to(\hbox{COMP}(\Pi_1,\emph{In},\emph{Out\/}) \lrar
\hbox{COMP}(\Pi_2,\emph{In},\emph{Out\/}))
\eeq{thmf}
\emph{is satisfied by all interpretations that are standard outside of~PH.}
\medskip

This theorem shows that the equivalence of tight programs may be established
by choosing a set \emph{Ax} of first-order axioms that are satisfied by
all interpretations standard outside of~\emph{PH}, and then deriving
formula~(\ref{thmf}) from \emph{Ax} in classical second-order logic.  Its
proof is given in Section~\ref{sec:proof}.

\section{Programs without private recursion}

In the special case when the io-programs
$(\Pi_1,\emph{PH},\emph{In},\emph{Out\/})$ and
$(\Pi_2,\emph{PH},\emph{In},\emph{Out\/})$ do not involve private recursion
\cite[Section~6.4]{fan20}, the use of second-order logic can be avoided in the
following way.  Represent the formula
COMP$(\Pi_1,\emph{In},\emph{Out\/})$ in the form
$$\exists {\bf P}\left(\bigwedge_i F_i({\bf P}) \land F'({\bf P})\right),$$
where $\bf P$ is a list
% $P_1,P_2,\dots$
of distinct predicate variables
corresponding to the
private symbols\footnote{A \emph{private symbol} of an io-program is
  a predicate symbol that occurs in its rules and is different from its
  input symbols and output symbols \cite[Section~5.1]{fan20}.}
$p_1,p_2,\dots$
of $(\Pi_1,\emph{PH},\emph{In},\emph{Out\/})$, and~$F_i({\bf P})$ is
the formula obtained from the completed definition \cite[Section~6.1]{fan20}
of~$p_i$ in~$\Pi_1$ by replacing each of $p_1,p_2,\dots$ by the
corresponding member of~$\bf P$. (The conjunctive members
of $F'({\bf P})$ correspond to the completed definitions of the output
symbols and to the constraints of~$\Pi_1$.)  Similarly, write
COMP$(\Pi_2,\emph{In},\emph{Out\/})$ as
\beq
\exists {\bf Q}\left(\bigwedge_j G_j({\bf Q}) \land G'({\bf Q})\right),
\eeq{comp2}
where $\bf Q$ is a list of distinct predicate variables corresponding to the
private symbols
of $(\Pi_2,\emph{PH},\emph{In},\emph{Out\/})$,
and the formulas $G_j({\bf Q})$ are obtained from the completed
definitions of these symbols in~$\Pi_2$ by replacing each of them
by the corresponding variable.  Take one half
\beq
\emph{Asm}\to(\hbox{COMP}(\Pi_1,\emph{In},\emph{Out\/}) \to
\hbox{COMP}(\Pi_2,\emph{In},\emph{Out\/}))
\eeq{thmf1}
of condition~(\ref{thmf}).
Since~$\Pi_2$ does not use private
recursion, formula~(\ref{comp2}) is equivalent to
$$\forall {\bf Q}\left(\bigwedge_j G_j({\bf Q}) \to G'({\bf Q})\right)$$
\cite[Theorem~3]{fan20}.  It follows that formula~(\ref{thmf1}) is
equivalent to
$$
\emph{Asm}\to
 \left(\exists {\bf P}\left(\bigwedge_i F_i({\bf P}) \land F'({\bf P})\right)
 \to\forall {\bf Q}\left(\bigwedge_j G_j({\bf Q}) \to G'({\bf Q})\right)\right)
$$
and consequently to
$$
\forall {\bf PQ}\left(\left(
    \emph{Asm}\land\bigwedge_i F_i({\bf P}) 
\land\bigwedge_j G_j({\bf Q}) \right) \to (F'({\bf P})\to G'({\bf Q}))\right)
$$
(with the bound variables in {\bf P}, {\bf Q} renamed, if necessary, to ensure
that they are pairwise disjoint).
Similarly, the second half
$$%\beq
\emph{Asm}\to(\hbox{COMP}(\Pi_2,\emph{In},\emph{Out\/}) \to
\hbox{COMP}(\Pi_1,\emph{In},\emph{Out\/}))
$$%\eeq{thmf12}
of condition~(\ref{thmf}) is equivalent to
$$
\forall {\bf PQ}\left(\left(
    \emph{Asm}\land\bigwedge_i F_i({\bf P}) 
\land\bigwedge_j G_j({\bf Q}) \right) \to (G'({\bf P})\to F'({\bf Q}))\right).
$$
Thus~(\ref{thmf}) can be rewritten as
$$
\forall {\bf PQ}\left(\left(
   \emph{Asm}\land\bigwedge_i F_i({\bf P}) 
   \land\bigwedge_j G_j({\bf Q}) \right) \to
          (F'({\bf P})\lrar G'({\bf Q}))\right).
$$
Finally, observe that
the derivability of this formula from \emph{Ax} is equivalent to the
derivability of the first-order formula
\beq
\left(
    \emph{Asm}\land\bigwedge_i F_i({\bf p})
\land\bigwedge_j G_j({\bf q}) \right) \to (F'({\bf p})\lrar G'({\bf q})),
\eeq{thmffo}
where {\bf p}, {\bf q} are lists of fresh predicate constants.

\section{Replacing programs by specifications}

We see that, under some conditions, proving the
equivalence of mini-\gringo\ programs can be reduced to deriving
formulas of form~(\ref{thmffo}).  This reduction is similar to the process of
proving program correctness that is implemented in \hbox{\anthem} \cite{fan20}.
The difference is that we talk here about the equivalence of a program to
another program, instead of the equivalence of a program to a specification.

The relationship between proving the equivalence of programs and the
operation of \anthem\ can be described also in a different way.
Define a specification~$\Sfr$ as follows:
\begin{itemize}
\item
the placeholders of~$\Sfr$ are the placeholders~\emph{PH} of user guide~(\ref{ug});
  \item
    the input symbols of~$\Sfr$ are the input symbols~\emph{In} of the user guide
    and the private symbols~{\bf p} of the io-program
$(\Pi_1,\emph{PH},\emph{In},\emph{Out\/})$;
\item
  the assumptions of~$\Sfr$ are the assumptions~\emph{Asm} of the user guide
  and the formulas~$F_i({\bf p})$;
\item
  the specs of~$\Sfr$ are the conjunctive terms of $F'({\bf p})$.
\end{itemize}
If we instruct \anthem\ to prove the claim
that the io-program $(\Pi_2,\emph{PH},\emph{In},\emph{Out})$
implements~$\Sfr$, then \anthem\ will look for a derivation of the formula
$$\left(
    \emph{Asm}\land\bigwedge_i F_i({\bf p})
\land\bigwedge_j G_j({\bf q}) \right) \to (G'({\bf q})\lrar F'({\bf p}))
$$
from \emph{Ax} \cite[Section~6.4]{fan20}.
This formula is equivalent to~(\ref{thmffo}).
Thus instructing \anthem\ to verify that
the io-program $(\Pi_1,\emph{PH},\emph{In},\emph{Out})$
implements the specification~$\Sfr$ amounts verifying the derivability
of~(\ref{thmffo}).

\section{Implementation}\label{sec:implementation}

[ Use the example from the introduction. ]

\section{Proof of Theorem \ref{thm1}}\label{sec:proof}

\begin{lem} \label{lem1}
In the assumptions of Definition~1, the external behavior of~$\Pi$ is
the collection of all sets~$\J$ of output atoms such that
$\I\cup\J$ is an io-model of the io-program
\emph{(}$\Pi$,PH,In,Out\emph{)}.
\end{lem}

\noindent{\bf Proof}$\;$
The external behavior of~$\Pi$ can be described as the
collection of all sets that can be represented as the intersection of
an io-model of ($\Pi$,\emph{PH},\emph{In},\emph{Out})
with the set of output atoms.  Every io-model of this io-program
is the union of~$\I$ with some
set~$\J$ of output atoms, and the intersection of this union with the
set of output atoms is~$\J$.

% Inputs of user guide~(\ref{ug}) can be characterized as the
% inputs $(v,\I)$ of io-program~(\ref{iop}) such that~$\I^\uparrow$
% satisfies~\emph{Asm}.

\medskip
\begin{lem}\label{lem2}\strut\cite[Theorem~2]{fan20} $\;$
  Let~$(v,\I)$ be an input for a tight
  io-program~\emph{(}$\Pi$,PH,In,Out\emph{)}.  For any
  set~$\J$ of output atoms, $\I\cup\J$ is an io-model of this io-program
  for the input~$(v,\I)$ iff the interpretation $v[(\I\cup\J)^\uparrow]$
  satisfies the program's completion.
\end{lem}

\noindent{\bf Proof of Theorem~\ref{thm1}}$\;$
Consider a user guide
(\emph{PH},\emph{In},\emph{Out},\emph{Asm}),
and take tight mini-\gringo\ programs $\Pi_1$,~$\Pi_2$ that do
not contain its input symbols in the heads of rules.
By Lemma~\ref{lem1}, the assertion that
$\Pi_1$ is equivalent to~$\Pi_2$  can be reformulated as follows:
\beq\ba c
\!\!\hbox{if $(v,\I)$ is an input for (\emph{PH},\emph{In},\emph{Out},\emph{Asm})
and~$\J$ is a set of output atoms then}\\
\I\cup\J \hbox{ is an io-model of ($\Pi_1$,\emph{PH},\emph{In},\emph{Out})}\\
\hbox{iff}\\
\hbox{it is an io-model of ($\Pi_2$,\emph{PH},\emph{In},\emph{Out}).}
\ea\eeq{p1}
Every input for (\emph{PH},\emph{In},\emph{Out},\emph{Asm}) is also an
input for each of the io-programs $(\Pi_1,\emph{In},\emph{Out\/})$,
$(\Pi_1,\emph{In},\emph{Out\/})$.  Consequently Lemma~\ref{lem2} allows us to
conclude that~(\ref{p1}) is equivalent to the condition
$$\ba c
\!\!\hbox{if $(v,\I)$ is an input for (\emph{PH},\emph{In},\emph{Out},\emph{Asm})
and~$\J$ is a set of output atoms then}\\
  v[(\I\cup\J)^\uparrow]\models
  \hbox{COMP}(\Pi_1,\emph{In},\emph{Out\/})\lrar
  \hbox{COMP}(\Pi_2,\emph{In},\emph{Out\/}).
\ea
$$
Thus~$\Pi_1$ is equivalent to~$\Pi_2$ iff for every set~$\I$ of input
atoms, every set~$\J$ of output atoms, and every function~$v$ that maps
the members of~\emph{PH} to precomputed terms that do not belong to~\emph{PH},
\beq \ba l
\hbox{if $v([\I^\uparrow])\models\emph{Asm}$}\\
\hbox{then
  $v[(\I\cup\J)^\uparrow]\models
  \hbox{COMP}(\Pi_1,\emph{In},\emph{Out\/})\lrar
  \hbox{COMP}(\Pi_2,\emph{In},\emph{Out\/})$.}
\ea\eeq{p2}
Furthermore, the condition $v([\I^\uparrow])\models\emph{Asm}$
is equivalent to \hbox{$v([(\I\cup\J)^\uparrow])\models\emph{Asm}$},
because the
assumptions \emph{Asm} do not contain output symbols.  Consequently~(\ref{p2})
can be rewritten as
$$  v[(\I\cup\J)^\uparrow]\models \emph{Asm}\to(
  \hbox{COMP}(\Pi_1,\emph{In},\emph{Out\/})\lrar
  \hbox{COMP}(\Pi_2,\emph{In},\emph{Out\/})).
$$
It remains to observe that an interpretation of~$\sigma(\emph{In})$ can be
represented in the form $v[(\I\cup\J)^\uparrow]$ iff it is standard
outside of~\emph{PH}.


\section{The language of assumptions}


\end{document}

User guides can be represented in the format of {\sc anthem} specification
files.  For example, the file
\begin{verbatim}
  input: a, b.
  assume: exists N (a = N) and exists N (b = N).
  assume: a > 1.
  output: prime/1.
\end{verbatim}
represents user guide~(\ref{ug}) with
The first two lines in that file can be written more concisely as
\begin{verbatim}
  input: a -> integer, b -> integer.
\end{verbatim}
Unlike general specification files, user guide files do not contain
\verb|spec| statements.

\section{External behavior and equivalence}

For any interpretation~$I$ of the signature~$\sigma(\emph{In})$
and any function~$v$ that maps
symbolic constants to precomputed terms, by $v[I]$ we denote the
interpretation obtained from~$I$ by reinterpreting every
constant~$c$ in the domain of~$v$ according to the formula
$$v[I](c)= I(v(c)).$$

An~\emph{input} for a user guide~\emph{UG} is a pair $(v,\I)$, where
\begin{itemize}
\item[(i)]
  $v$ is a function that maps the placeholders of~\emph{UG} to precomputed
terms that are not placeholders, and
\item[(ii)]
  $\I$ is a subset of the set of input atoms of~\emph{UG} such that
  the interpretation~$v[\I^\uparrow]$ satisfies the assumptions
  of~\emph{UG}.\footnote{The meaning of~$\I^\uparrow$ is defined in the
    appendix.}
\end{itemize}
